String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-05-16 2023-1-19/sections/algorithms.tex
references
more revises

\begin{definition}[Nonemptiness of CEFA w.r.t. an LIA formula\footnote{The decision problem was called satisfiability of an LIA formula w.r.t. CEFA in \cite{atva2020}.}, abbreviated as $\cefadec$]
Suppose that $x_1, \cdots, x_n$ are string variables, $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are nonempty sets of CEFA over the alphabet $\Sigma$ with $\Lambda_{x_i} = \{\aut_{i, 1}, \cdots, \aut_{i, l_i}\}$ for every $i \in [n]$, and the sets of registers $R_{\aut_{1, 1}}, \cdots, R_{\aut_{1, l_1}}, \cdots, R_{\aut_{n, 1}}, \cdots, R_{\aut_{n, l_n}}$ are mutually disjoint, moreover, $\varphi$ is an LIA formula whose free variables are from $R' = \bigcup \limits_{i \in [n], j \in [l_i]} R_{\aut_{i, j}}$. Then  the CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are said to be nonempty w.r.t. $\varphi$ if there are an assignment $\theta: \{x_1, \cdots, x_n\} \rightarrow \Sigma^*$ and vectors $\myvec{v_{i,j}}$ for $i \in [n], j \in [l_i]$
%$\myvec{v_{1,1}}, \cdots, \myvec{v_{1, l_1}}, \cdots, \myvec{v_{n, 1}}, \cdots, \myvec{v_{n, l_n}}$ 
such that $(\theta(x_i); \myvec{v_{i,j}}) \in \Lang(\aut_{i, j})$ for every $i \in [n], j \in [l_i]$,
% \in \aut_{i,1}(\eta(x_1)), \cdots, \myvec{v_{i, l_i}} \in \aut_{i,l_i}(\eta(x_i))$ for every $i \in [n]$  
%\cdots, \myvec{v_{n,1}} \in \aut_{n,1}(w_n), \cdots, \myvec{v_{n, l_n}} \in \aut_{n, l_n}(w_n)$ 
and 
$\varphi[(\myvec{v_{i,j}}/R_{\aut_{i,j}})_{i \in [n], j \in [l_i]}]$ is true. 
%$\varphi[\myvec{v_{1,1}}/R_{\aut_{1,1}}, \cdots, \myvec{v_{n, l_n}}/R_{\aut_{n, l_n}}]$ is true. 
\end{definition}

\begin{proposition}[\cite{atva2020}]\label{prop-cefadec}
$\cefadec$ is PSPACE-complete. 
\end{proposition}
The decision procedure in \cite{atva2020} was only used to prove the upper bound in Proposition~\ref{prop-cefadec} and not implemented as a matter of fact. Instead, the symbolic model checker nuXmv was used to solve $\cefadec$ in \cite{atva2020}. We do not rely on nuXmv in this work and shall propose an efficient algorithm for solving $\cefadec$ in Section~\ref{subsec:cefadec}.

@@

\vspace{-2mm}
\subsection{From $\reclsat$ to $\cefadec$} \label{subsec:regex2cefa}

@@

For non-register-representable regular expressions, we first transform them into register-representable regular expressions by unfolding all the non-register-representable occurrences of  counting operators, then utilize the aforementioned procedure to construct CEFA. For instance, $(a^{\{2, 6\}} \concat b^*)^{\{2,\infty\}}$ is transformed into $(aa(\varepsilon + a + aa + aaa+aaaa) \concat b^*)^{\{2, \infty\}}$. 
The unfoldings of the inner counting operators of non-register-representable regexes incur an exponential blow-up in the worst case. Nevertheless, those regexes occupy only 5\% of the 40,628 regexes that are collected from the practice (see Section~\ref{sec:bench}). Moreover, the unfoldings are partial in the sense that the outmost counting operators are not unfolded. It turns out that our approach can solve almost all of them, except 106.   
%$(a^{\{2, 6\}} \concat b^*) \setminus (a + b)^{\{4, \infty\}}$ is transformed into $(aa(\varepsilon + a + aa + aaa + aaaa) \concat b^*) \setminus (a+b)(a+b)(a+b)(a+b)(a+b)^*$.

@@

\begin{observation}\label{obs-cefa-output}
CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$ iff there are $\myvec{v_{1}} \in \cefaout(\cB_1), \cdots, \myvec{v_{n}} \in \cefaout(\cB_n)$ such that $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true.
\end{observation}

@@

Let $\aut = (R, Q, \Sigma, \delta, I, F, \alpha)$ be a CEFA and $\costset_\aut = \{\myvec{v} \mid \exists q, a, q'.\ (q, a, q', \myvec{v}) \in \delta\}$. 
Moreover, let $\cU_\aut = (Q, \costset_\aut, \delta', I, F)$ be an NFA over the alphabet $\costset_\aut$ that is obtained from $\aut$ by dropping the accepting condition and ignoring the characters, that is, $\delta'$ comprises the tuples $(q, \myvec{v}, q')$ such that $(q, a, q', \myvec{v}) \in \delta$ for some $a \in \Sigma$.
%Then $\cefaout(\aut) = \parikh(\NFA')$.
%
\begin{observation}\label{obs-cefa-nfa}
For each CEFA $\aut = (R, Q, \Sigma, \delta, I, F, \alpha)$, 
%
$$\cefaout(\aut) = \left\{\sum \limits_{\myvec{v} \in \costset_\aut} \eta(\myvec{v}) \myvec{v} \  \big{|}\ \eta \in \parikh(\cU_\aut) \mbox{ and }  \alpha\left[ \sum \limits_{\myvec{v} \in \costset_\aut} \eta(\myvec{v}) \myvec{v} \big{/} R \right] \mbox{ is true}\right\}.$$
%
\end{observation}

>>>>

\vspace{-3mm}
\begin{definition}[Nonemptiness of CEFA w.r.t. an LIA formula\footnote{The decision problem was called satisfiability of an LIA formula w.r.t. CEFA in \cite{atva2020}.}, abbreviated as $\cefadec$]
Suppose that $x_1, \cdots, x_n$ are string variables, $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are nonempty sets of CEFA over the alphabet $\Sigma$ with $\Lambda_{x_i} = \{\aut_{i, 1}, \cdots, \aut_{i, l_i}\}$ for every $i \in [n]$, and the sets of registers $R_{\aut_{1, 1}}, \cdots, R_{\aut_{1, l_1}}, \cdots, R_{\aut_{n, 1}}, \cdots, R_{\aut_{n, l_n}}$ are mutually disjoint, moreover, $\varphi$ is an LIA formula whose free variables are from $R' = \bigcup \limits_{i \in [n], j \in [l_i]} R_{\aut_{i, j}}$. Then  the CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are said to be nonempty w.r.t. $\varphi$ if there are an assignment $\theta: \{x_1, \cdots, x_n\} \rightarrow \Sigma^*$ and vectors $\myvec{v_{i,j}}$ for $i \in [n], j \in [l_i]$
%$\myvec{v_{1,1}}, \cdots, \myvec{v_{1, l_1}}, \cdots, \myvec{v_{n, 1}}, \cdots, \myvec{v_{n, l_n}}$ 
such that $(\theta(x_i); \myvec{v_{i,j}}) \in \Lang(\aut_{i, j})$ for every $i \in [n], j \in [l_i]$,
% \in \aut_{i,1}(\eta(x_1)), \cdots, \myvec{v_{i, l_i}} \in \aut_{i,l_i}(\eta(x_i))$ for every $i \in [n]$  
%\cdots, \myvec{v_{n,1}} \in \aut_{n,1}(w_n), \cdots, \myvec{v_{n, l_n}} \in \aut_{n, l_n}(w_n)$ 
and 
$\varphi[(\myvec{v_{i,j}}/R_{\aut_{i,j}})_{i \in [n], j \in [l_i]}]$ is true. 
%$\varphi[\myvec{v_{1,1}}/R_{\aut_{1,1}}, \cdots, \myvec{v_{n, l_n}}/R_{\aut_{n, l_n}}]$ is true. 
\end{definition}

\vspace{-4mm}
\begin{proposition}[\cite{atva2020}]\label{prop-cefadec}
$\cefadec$ is PSPACE-complete. 
\end{proposition}
The decision procedure in \cite{atva2020} was only used to prove the upper bound in Proposition~\ref{prop-cefadec} and not implemented as a matter of fact. Instead, the symbolic model checker nuXmv was used to solve $\cefadec$ in \cite{atva2020}. We do not rely on nuXmv in this work and shall propose an efficient algorithm for solving $\cefadec$ in Section~\ref{subsec:cefadec}.

@@

\vspace{-3mm}
\subsection{From $\reclsat$ to $\cefadec$} \label{subsec:regex2cefa}

@@

For non-register-representable regular expressions, we first transform them into register-representable regular expressions by unfolding all the non-register-representable occurrences of  counting operators, then utilize the aforementioned procedure to construct CEFA. For instance, $(a^{\{2, 6\}} \concat b^*)^{\{2,\infty\}}$ is transformed into $(aa(\varepsilon + a + aa + aaa+aaaa) \concat b^*)^{\{2, \infty\}}$. 
The unfoldings of the inner counting operators of non-register-representable regexes incur an exponential blow-up in the worst case. Nevertheless, those regexes occupy only 5\% of the 48,834 regexes that are collected from the practice (see Section~\ref{sec:bench}). Moreover, the unfoldings are partial in the sense that the outmost counting operators are not unfolded. It turns out that our approach can solve almost all of them, except 145.   
%$(a^{\{2, 6\}} \concat b^*) \setminus (a + b)^{\{4, \infty\}}$ is transformed into $(aa(\varepsilon + a + aa + aaa + aaaa) \concat b^*) \setminus (a+b)(a+b)(a+b)(a+b)(a+b)^*$.

@@

\vspace{-2mm}
\begin{observation}\label{obs-cefa-output}
CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$ iff there are $\myvec{v_{1}} \in \cefaout(\cB_1), \cdots, \myvec{v_{n}} \in \cefaout(\cB_n)$ such that $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true.
\end{observation}

@@

Let $\aut = (R, Q, \Sigma, \delta, I, F, \alpha)$ be a CEFA and $\costset_\aut = \{\myvec{v} \mid \exists q, a, q'.\ (q, a, q', \myvec{v}) \in \delta\}$. 
Moreover, let $\cU_\aut = (Q, \costset_\aut, \delta', I, F)$ be an NFA over the alphabet $\costset_\aut$ that is obtained from $\aut$ by dropping the accepting condition and ignoring the characters, that is, $\delta'$ comprises the tuples $(q, \myvec{v}, q')$ such that $(q, a, q', \myvec{v}) \in \delta$ for some $a \in \Sigma$.
%Then $\cefaout(\aut) = \parikh(\NFA')$.
%
\vspace{-2mm}
\begin{observation}\label{obs-cefa-nfa}
For each CEFA $\aut = (R, Q, \Sigma, \delta, I, F, \alpha)$, 
%
$$\cefaout(\aut) = \left\{\sum \limits_{\myvec{v} \in \costset_\aut} \eta(\myvec{v}) \myvec{v} \  \big{|}\ \eta \in \parikh(\cU_\aut) \mbox{ and }  \alpha\left[ \sum \limits_{\myvec{v} \in \costset_\aut} \eta(\myvec{v}) \myvec{v} \big{/} R \right] \mbox{ is true}\right\}.$$
%
\end{observation}
